---
title: "Problem Set Appication 3"
author: "Sabrina Li"
date: "28/05/2021"
output: 
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
urlcolor: blue
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
Sys.setlocale(locale = 'USA')
```

<!-- .Rmd files use  markdown, a text mark up language, to provide formating.--> 
<!--Text include within these strange arrows are comments and will not show up when you knit-->

# Front matter
This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **__**

Add your collaborators: **__**

Late coins used this pset: 0. Late coins left: X. 
<!--You may use up to two for a given assignment.)-->

# Problems

## Load the data and first glimpse



### 1
```{r}
df = read.csv('gun-violence-data_01-2013_03-2018.csv')

dim(df)
```


### 2


for each recored, we have location, number and background of participants, the guns useed, and the result.
```{r}

```

### 3

```{r}
colnames(df)[20: 26]
```


## Time ralated trends of gun violence


### 1

```{r}
df$date = ymd(df$date)
```


### 2

```{r}
df$date_extr = df$date %>% str_extract(., '[0-9]{4}-[0-9]{1,2}-[0-9]{1,2}')
df$date_d = ymd(df$date_extr)

df$year = df$date_d %>% year()

df %>% filter(year %>% is.na() == F) %>% group_by(year) %>%
  summarise(count = n()) %>% ggplot(aes(x = year, y = count)) +
  geom_line() +  geom_text(aes(label = count), 
                           vjust = "inward", hjust = "inward") + 
  labs(title = 'time series: number of incidents' )

```


Yes, for year 2018, we only have data of 3 months.

### 3

```{r}
df$month = df$date_d %>% month() 

df %>% group_by(year) %>% summarise(n_distinct(month))

df %>% filter(month %>% is.na() == F & year < 2018)  %>% ggplot(aes(x = factor(month))) +
  geom_bar() +  geom_text(stat='count', aes(label=..count..), vjust=-0.5) +  labs(title = ' number of incidents by months' , x = 'month')

```
Summer time, i.e, July and August, there are more cases, but it is not very siginificant.



### 4

```{r}
df$weekday = df$date_d %>% weekdays() 


df %>% filter(weekday %>% is.na() == F & year < 2018)  %>% ggplot(aes(x = factor(weekday))) +
  geom_bar() +  geom_text(stat='count', aes(label=..count..), vjust=-0.5) +  labs(title = ' number of incidents by weekdays' , x = 'weekday')


```


Sunday has most cases.


### 5


the number of incidents increasing by year, comparing to other months, July and August have more incidents, and for weekdays, Sunday has more cases, and Thursday has the least ones.



## Characteristics of Gun violence incidents

### 1

```{r}
df$n_guns_involved %>% mean(na.rm =T)
```
This is the average number of guns involved. We have not consider the NA value.

```{r}
df$n_guns_involved ->ng
ng[ng %>% is.na()] = 0
mean(ng)
```
If NA means 0, the figure is above.


### 2


```{r}
df$gun_type[df$gun_type %in% c("", "0::Unknown") == F] -> guntypes

(guntypes %>% str_split('\\|\\|', simplify = T) %>% str_split('::', simplify = T))[,2] -> gun_types

gun_types[gun_types %in% c("", 'Unknown') ==F] -> gun_types

data.frame(types = gun_types) %>% ggplot(aes(y =types)) + geom_bar() + labs(title = 'gun types count')


```



Handguns are most used.


### 3

It is just the characteristics of an incident, an incident coudl have multiple attributes.


### 4

```{r}
df$incident_characteristics %>% str_split('\\|', simplify = F) -> incd_char
incd_char %>% unlist() -> incd_char

ind_df = data.frame(types = incd_char[incd_char != ''] ) %>% group_by(types) %>% summarise(count = n()) %>% arrange(desc(count)) %>% head(20)
ind_df %>% 
  ggplot(., aes(y=types, x=count))+
  geom_bar(stat='identity') + labs(title = 'incident_characteristics', subtitle = '20 most common')


```


## Suspects Characteristics


### 1

There were often more than one people involved in the cases, and their identitis are keys and for each variables they have different values.



### 2


```{r}
df$participant_type %>% str_split('\\|',simplify = F) -> part_types


count_vic = map(part_types, function(x){
  
  nv = x %>% str_extract_all('Victim') %>% unlist() %>% length()
  
  return(nv)
  
  
})


count_vic %>% unlist() %>% mean() # the numebr of vicitms



count_susp = map(part_types, function(x){
  
  ns = x %>% str_extract_all('Suspect') %>% unlist() %>% length()
  
  return(ns)
  
  
})

count_susp %>% unlist() %>% mean() # the number of suspects
```


The average number we have are listed above.



### 3

```{r}
df %>% select(incident_id,
              participant_age, participant_gender) -> sub_df
head(sub_df)

```



### 4

```{r}
sub_df$nest_age = sub_df$participant_age %>% str_split('\\|\\|', simplify= F)
tmp = sub_df %>% unnest(c(nest_age)) 
tmp$nest_age %>% str_split("::", simplify = T) -> tmp_age

summary(tmp_age[,2] %>% as.numeric())



data.frame(age = tmp_age[,2] %>% as.numeric()) %>% ggplot(aes(age)) + geom_density() + labs(title = 'distirbution of age')

```

we do have records in which  ages larger than 100 are contained.


### 5


```{r}
sub_df$nest_gender = sub_df$participant_gender %>% str_split('\\|\\|', simplify= F)
tmp = sub_df %>% unnest(c(nest_gender)) 
tmp$nest_gender %>% str_split("::", simplify = T) -> tmp_gender

gender_dt = tmp_gender[,2] 
gender_dt[gender_dt %in% c('Male', 'Female') ] %>% table() %>% prop.table()

```

over 87 percentages are males.


### 6


```{r}
df$participant_status %>% str_split('\\|',simplify = F) -> part_status


part_status %>% unlist() %>% unique() %>% str_extract('[a-zA-Z]+') %>% setdiff(NA)


```
There are four.




### 7

```{r}
count_arrest = map(part_status, function(x){
  
  ns = x %>% str_extract_all('Arrested') %>% unlist() %>% length()
  
  return(ns)
  
  
}) %>% unlist() %>% sum() -> n_arrest

count_susp %>% unlist() %>% sum() -> n_susp

n_arrest / n_susp
```

49.85% were arrested.

## Geographic variation


### 1

```{r}
library(tidycensus)
census_api_key("f51cd09bd6d0f8cdc620d3c64a5f7893c2efebe1")
options(tigris_use_cache = TRUE)
pop_df <- get_acs(geography = "state", 
                  variables = "B01003_001", 
                  year = 2017,
                  geometry = TRUE) 

df17 = df[df$year == 2017,]

df17 %>% ggplot(aes(y = state)) + geom_bar() + labs(title = 'case number in 2017')

```

Illinois has it.

### 2


```{r}
count17 = df17 %>% group_by(state) %>% summarise(count = n()) 
colnames(count17)[1] = 'NAME'
map_df = left_join( pop_df,count17)
map_df$density = map_df$count / map_df$estimate


map_df %>% ggplot(aes(y = NAME, x = density)) + geom_bar(stat = 'identity')+ labs(title = 'case rate in 2017')

```


District of Columbia has it.

### 3


```{r}
map_df  %>% ggplot(aes(fill = density)) +
  geom_sf() +
  labs(title = "incident rate  by state",
       caption = "Year 2017")
```



## Mass shootings


### 1

```{r}
df$participant_type %>% str_split('\\|',simplify = F) -> part_types

count_vic = map(part_types, function(x){
  
  nv = x %>% str_extract_all('Victim') %>% unlist() %>% length()
  
  return(nv)
  
  
})
df$count_vic = count_vic %>% unlist() 
df %>% arrange(count_vic) %>% tail(15) ->id15
```


### 2

```{r}
id15 %>% ggplot(aes(longitude, latitude)) +
  borders("state") +
  geom_point(aes(size =count_vic )) +
  coord_quickmap() +
  labs(title = 'top 15 incidents by number of victims')

```


### 3



```{r}
id15$nest_age = id15$participant_age %>% str_split('\\|\\|', simplify= F)
tmp = id15 %>% unnest(c(nest_age)) 
tmp$nest_age %>% str_split("::", simplify = T) -> tmp_age

summary(tmp_age[,2] %>% as.numeric())



data.frame(age = tmp_age[,2] %>% as.numeric()) %>% ggplot(aes(age)) + geom_density() + labs(title = 'distirbution of age')

```

The ages that in mass shooting cases are younger.


```{r}
id15$nest_gender = id15$participant_gender %>% str_split('\\|\\|', simplify= F)
tmp = id15 %>% unnest(c(nest_gender)) 
tmp$nest_gender %>% str_split("::", simplify = T) -> tmp_gender

gender_dt = tmp_gender[,2] 
gender_dt[gender_dt %in% c('Male', 'Female') ] %>% table() %>% prop.table()


```

The proportion of males is less.


```{r}
id15$participant_status %>% str_split('\\|',simplify = F) -> part_status


part_status %>% unlist() %>% unique() %>% str_extract('[a-zA-Z]+') %>% setdiff(NA)




id15$participant_type %>% str_split('\\|',simplify = F) -> part_types


count_vic = map(part_types, function(x){
  
  nv = x %>% str_extract_all('Victim') %>% unlist() %>% length()
  
  return(nv)
  
  
})


count_vic %>% unlist() %>% mean()



count_susp = map(part_types, function(x){
  
  ns = x %>% str_extract_all('Suspect') %>% unlist() %>% length()
  
  return(ns)
  
  
})

count_susp %>% unlist() %>% mean()

```
The number of people involved in mass shootings are much more.


```{r}
count_susp %>% unlist() %>% sum() -> n_susp



id15$participant_status %>% str_split('\\|',simplify = F) -> part_status
count_arrest = map(part_status, function(x){
  
  ns = x %>% str_extract_all('Arrested') %>% unlist() %>% length()
  
  return(ns)
  
  
}) %>% unlist() %>% sum() -> n_arrest

n_arrest / n_susp
```
The arrest rate in such cases are higher.



